---
title: "Lead IQ Exploratory Data Analysis"
author: "Rachel Austin"
date: "2024-10-07"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE,message = FALSE}
lead_df <- read.csv("/Users/austinra/Documents/GitHub/Austin_Project_01/DataRaw/lead-iq-01.csv")
```

<br>
```{r, echo = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
ggplot(lead_df,aes(Smelter, IQ, fill = Smelter)) + geom_boxplot()+coord_flip() + labs(title = "Distance from Lead Smelter vs IQ Levels", caption = c('Near (less than one mile), Far (greater than one mile)'), x= 'Distance from Smelter')+ theme_minimal()+scale_fill_brewer(palette = "Spectral")+scale_fill_discrete(labels = c('Far from Lead Smelter', 'Near Lead Smelter'))+guides(fill=guide_legend(reverse=TRUE))

```
<br>
<br>
<br>
```{r, echo = FALSE, message = FALSE}
# aggregated table
library(knitr)
library(kableExtra)

nearsmelter <- filter(lead_df, Smelter == 'Near')
farsmelter <- filter(lead_df, Smelter == 'Far')
IQR_nearsmelter <- as.matrix(summary(nearsmelter$IQ))
colnames(IQR_nearsmelter) <- c('Less than 1 mile from Lead Smelter')
IQR_farsmelter <- as.matrix(summary(farsmelter$IQ))
colnames(IQR_farsmelter) <- c('More than 1 mile from Lead Smelter')

IQ_IQR <- cbind(IQR_nearsmelter, IQR_farsmelter)
IQ_IQR <- round(IQ_IQR, 2)
kable(IQ_IQR, caption = 'IQ Summary Table')

```
<br>

The boxplot is difficult to interpret as there is a very large IQ value in the Far from Lead Smelter group. The table shows us a more detailed summary of the Near Smelter and Far Smelter groups IQ values. The two groups have similar interquartile ranges, but the mean of the Far Smelter group IQ is higher than the Near Smelter group IQ. This might be driven by our extreme maximum value in this group of 999. 
<br>
<br>
Less than 1 mile from Lead Smelter Mean IQ - `r round(mean(nearsmelter$IQ),2)`
<br>
More than 1 mile from Lead Smelter Mean IQ - `r round(mean(farsmelter$IQ),2)`
<br>
<br>
Below is a summary of the IQ levels of all people in the data set, regardless of distance from lead smelter.
```{r}
summary(lead_df$IQ)
```
